\vspace{-2cm} \LARGE Divergence-Free Smoothed Particle Hydrodynamics \\ \large Advanced Simulation and Visualization of Fluids in Computer Graphics

### 2017-01-12 report/paper.tex
small changes in the report, added look-up-table to future work

\subsection{Navier-Stokes}
    Navier-Stokes equations describes incompressible fluids and is therefore suitable for SPH simulations.
    The equations expressed in Lagrangian coordinates can be found in Equation~\ref{eq:nsDensity} and~\ref{eq:nsVelocity}.

@@

    In the equations $\frac{D\rho}{Dt}$ and $\frac{D\textbf{v}}{Dt}$ denotes the derivate of the density $\rho$ and the velocity $\textbf{v}$ respectively.
    The pressure, kinematic viscosity and body forces are denoted by $p,v$ and $\textbf{f}$ respectively.
    The second part of Equation~\ref{eq:nsDensity} $\nabla \cdot \textbf{v} = 0$ induce that the velocity field is free from divergence.
    From the continuity equation $\frac{D\rho}{Dt} = -\rho \nabla \cdot \textbf{v}$ and the divergence-free condition it follows that the partial derivate of the density with respect to the time is zero, which implies the equivalence of the equations in Equation~\ref{eq:nsDensity}.
    From this is follows that the density, in theory, must stay constant over time which is the implication of an incompressible fluid.
    In practice the divergence-free condition is not sufficient to guarantee incompressibility in simulations.
    The numerical time integrations will contain numerical errors which will cause density deviations due to the volume compressions.
    To avoid this a second condition $\rho - \rho_0 = 0$ called \textit{constant density condition} must be fullfilled~\cite{bender}.
    Section~\ref{divergenceSolver} and~\ref{densitySolver} give a deeper description of the solvers used in the project.

@@

% CFL condition, explain why its necessary and how its assured that each rendered frame has the same time step.
In order to assure a stable simulation it is necessary to adapt the time step depending on the motion.
The quicker the particles move the smaller time steps are taken such that the numerical 
error of the explicit time integration is negligible.
For this is a condition proposed by 
Courant$-$Friedrichs$-$Lewy used often refered as the CFL-condition show in equation \ref{eq:cfl}.
The time step $\Delta t$ is equal to a fraction of the particle's diameter $d$ divided by 
the maximum velocity $\vect{V_{max}}$ in the scene.

\begin{equation} \label{eq:cfl}
\Delta t \leq \frac{0.4 \cdot d}{\vect{V_{max}}}
\end{equation}

Each rendered frame however have a constant time step in order to ensure no speed change in the playback of the rendered video.
This is decided from the user side as a desired fps parameter.
This means that the simulation loop can iterate multiple times within each frame adaptively depending on the maximum velocity.

\subsection{Neighbourhood search} \label{section:neighbour}
    Since SPH only considers a finite amount of neighbouring particles it is important to keep track of every particles neighbours.
    Searching through all particles for neighbours within the cutoff distance $h$ for every particle is inefficient and takes $\mathcal{O}({N^2})$ time.
    The cutoff distance $h$ is the kernels smoothing radius.
    To make this faster a cell list was implemented.
    A cell list is a data structure that is divided into cells that have a length larger or equals to the cutoff distance $h$.
    Each cell spans on a surface in space.
    A particle belongs to a cell if it is inside the space the cell occupy.
    When finding the neighbour of particle $i$, only the neighbouring cells have to be searched for particles within the cutoff distance, see figure~\ref{fig:cellList}.

@@

    The cell list is implemented by using a four dimensional vector where the first three dimensions are for the x, y and z coordinates and the fourth dimension is for storing the particles belonging to the cell.
    The amount of cells are decided by dividing the scene into cells of length $h$.
    The particles are then assigned to a cell by finding the cell's coordinates that it belongs to according to Equation ~\ref{eq:assignCell}, where $\vect{x}$ is the position of the particle and $\vect{s}_min$ is the lowest position of the scene space.
    If the particle moves out of its cell it is reassigned to the new cell it is inside.

@@

    \begin{equation} \label{eq:predDensity}
        \resizebox{0.85\linewidth}{!}{$
        \begin{matrix}
        \rho_i^* = \rho_i + \Delta t \frac{D\rho_i}{Dt} \\ 
        \\
        \Leftrightarrow \\ 
        \\
        \rho_i^* = \rho_i + \Delta t + \sum_j m_j 
        (\textbf{v}_i^* - \textbf{v}_j^*) \nabla \vect{W}_{ij}
        \end{matrix}
        $}
    \end{equation}

@@

    \begin{equation} \label{eq:velocity}
        \resizebox{0.85\linewidth}{!}{$
        \vect{v_i^*}(t) = \vect{v_i^*}(t) - \Delta t \sum_j m_j (\frac{\kappa_i^v}{\rho_i} + \frac{\kappa_j^v}{\rho_i}) \nabla \vect{W}_{ij}
        $}
    \end{equation}

@@

    The stiffness parameters needs to be determined iteratively since neighboring particles depend on each other.
    After the symmetric pressure force has been applied to the new velocity should the density change in Equation \ref{eq:dpdt} be equal or close to zero.
    If the error is too large the process will be iterated until desired result have been achieved.

@@

An isosurface was also introduced in the middle of the scene in order to simulate collisions with arbitrary objects.
This was done in an implicit representation of geometric shapes, as function $f(x,y,z)$.
Given an isovalue $C = f(\vect{x})$, point $\vect{x}$ can be classified as follows:

@@

To include all possible implicit geometries can a quadratic function be described as:

@@

or on matrix form:

@@

Since the quadric surface is known analytically can the differentiation be applied to the quadric directly to get the gradient of the shape in position $\vect{x}$.
This gives an efficient and analytical expression with a coefficient matrix $Q$, which is used to help determine the new velocity direction of a particle after colliding with the surface.

@@

>>>>

\subsection{Navier-Stokes}
    Navier-Stokes equations describe incompressible fluids and are therefore suitable for SPH simulations.
    The equations expressed in Lagrangian coordinates can be found in Equation~\ref{eq:nsDensity} and~\ref{eq:nsVelocity}.

@@

    In the equations $\frac{D\rho}{Dt}$ and $\frac{D\textbf{v}}{Dt}$ denotes the derivate of the density $\rho$ and the velocity $\textbf{v}$ respectively.
    The pressure, kinematic viscosity and body forces are denoted by $p,v$ and $\textbf{f}$ respectively.
    The second part of Equation~\ref{eq:nsDensity} $\nabla \cdot \textbf{v} = 0$ induce that the velocity field is free from divergence.
    From the continuity equation $\frac{D\rho}{Dt} = -\rho \nabla \cdot \textbf{v}$ and the divergence-free condition it follows that the partial derivate of the density with respect to the time is zero, which implies the equivalence of the equations in Equation~\ref{eq:nsDensity}.
    From this is follows that the density, in theory, must stay constant over time which is the implication of an incompressible fluid.
    In practice the divergence-free condition is not sufficient to guarantee incompressibility in simulations.
    The numerical time integrations will contain numerical errors which will cause density deviations due to the volume compressions.
    To avoid this a second condition $\rho - \rho_0 = 0$ called \textit{constant density condition} must be fullfilled~\cite{bender}.
    Section~\ref{densitySolver} and~\ref{divergenceSolver} give a deeper description of the solvers used in the project.

@@

% CFL condition, explain why its necessary and how its assured that each rendered frame has the same time step.
In order to assure a stable simulation it is necessary to adapt the time step depending on the motion.
The quicker the particles move the smaller time steps are taken such that the numerical 
error of the explicit time integration is negligible.
A condition was proposed by 
Courant$-$Friedrichs$-$Lewy, often refered to as the CFL-condition, which is shown in Equation \ref{eq:cfl}.
The time step $\Delta t$ is equal to a fraction of the particle's diameter $d$ divided by 
the maximum velocity $\vect{V}_{max}$ in the scene.

\begin{equation} \label{eq:cfl}
\Delta t \leq \frac{0.4 \cdot d}{\vect{V}_{max}}
\end{equation}

Each rendered frame however have a constant time step in order to ensure no speed change in the playback of the rendered video.
This is decided from the user side as a desired fps, \textit{frames per second}, parameter.
This means that the simulation loop can iterate multiple times within each frame adaptively depending on the maximum velocity.

\subsection{Neighbourhood search} \label{section:neighbour}
    Since SPH only considers a finite amount of neighbouring particles it is important to keep track of every particles' neighbours.
    Searching through all particles for neighbours within the cutoff distance $h$ for every particle is inefficient and takes $\mathcal{O}({N^2})$ time.
    The cutoff distance $h$ is the kernels' smoothing radius.
    To make this faster a cell list was implemented.
    A cell list is a data structure that is divided into cells that have a length larger or equals to the cutoff distance $h$.
    Each cell spans on a surface in space.
    A particle belongs to a cell if it is inside the space the cell occupy.
    When finding the neighbour of particle $i$, only the neighbouring cells have to be searched for particles within the cutoff distance, see Figure~\ref{fig:cellList}.

@@

    The cell list is implemented by using a four dimensional vector where the first three dimensions are for the $x$, $y$ and $z$ coordinates and the fourth dimension is for storing the particles belonging to the cell.
    The amount of cells are decided by dividing the scene into cells of length $h$.
    The particles are then assigned to a cell by finding the cell's coordinates that it belongs to according to Equation ~\ref{eq:assignCell}, where $\vect{x}$ is the position of the particle and $\vect{s}_min$ is the lowest position of the scene space.
    If the particle moves out of its cell it is reassigned to the new cell it is inside.

@@

    \begin{equation} \label{eq:predDensity}
        \resizebox{0.85\linewidth}{!}{$
        \begin{matrix}
        \rho_i^* = \rho_i + \Delta t \frac{D\rho_i}{Dt} \\ 
        \\
        \Leftrightarrow \\ 
        \\
        \rho_i^* = \rho_i + \Delta t \sum_j m_j 
        (\textbf{v}_{i}^* - \textbf{v}_j^*) \nabla \vect{W}_{ij}
        \end{matrix}
        $}
    \end{equation}

@@

    \begin{equation} \label{eq:velocity}
        \resizebox{0.85\linewidth}{!}{$
        \vect{v_{i+1}^*}(t) = \vect{v_i^*}(t) - \Delta t \sum_j m_j (\frac{\kappa_i^v}{\rho_i} + \frac{\kappa_j^v}{\rho_i}) \nabla \vect{W}_{ij}
        $}
    \end{equation}

@@

    The stiffness parameters need to be determined iteratively since neighboring particles depend on each other.
    After the symmetric pressure force has been applied to the new velocity should the density change in Equation \ref{eq:dpdt} be equal or close to zero.
    If the error is too large the process will be iterated until desired result have been achieved.

@@

An isosurface was also introduced in the middle of the scene in order to simulate collisions with arbitrary objects.
This was done as an implicit representation of geometric shapes, described as a function $f(x,y,z)$.
Given an isovalue $C = f(\vect{x})$, point $\vect{x}$ can be classified as follows:

@@

All possible implicit geomitries can be descibed by a quadratic function, which is found in Equation~\ref{eqn:quadratic}, and in matrix form in Equation~\ref{eqn:quadricmatrixform}.

@@

@@

Since the quadric surface is known analytically the differentiation can be applied to the quadric directly to get the gradient of the shape in position $\vect{x}$.
This gives an efficient and analytical expression with a coefficient matrix $Q$, which is used to help determine the new velocity direction of a particle after colliding with the surface.

@@

    % Look up table for kernel values
    According to Bender et al.~\cite{bender} the kernel values could easily be precomputed and used together with a look up-table in order to speed up the calculations. This will result in a 30 $\%$ faster application.  This could ether be done once when the iteration loop begins and the positions for the particles are determined. Another approach would be to compute a discrete representation of the kernel function before the iteration loop and interpolate between to kernel values in the simulation loop in order to find a kernel value for a specific particle.
